Identifying Microlensing Events in Large, Non-Uniformly Sampled Surveys: The Case of the Palomar Transient Factory

### 2012-08-28 text/paper/ptf_microlensing_paper.tex
text updates and new referenceS

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as supernova searches, transiting exoplanet discoveries, or stellar variability studies, which set the cadence with which individual fields get re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to an extremely non-uniform sampling gradient over the survey footprint of nearly 20,000 deg$^2$: while the typical 7.26~deg$^2$ PTF field has been imaged 15 times, \apwsim1000~deg$^2$ of the survey has been observed more than 150 times. We use the existing PTF data to study the trade-off between a large survey footprint and irregular sampling when searching for microlensing events, and to examine the probability that such events can be recovered in these data. We conduct Monte Carlo simulations to evaluate our detection efficiency in a hypothetical survey field as a function of both the baseline and number of observations. We also apply variability statistics to systematically differentiate between periodic, transient, and flat light curves. Preliminary results suggest that both recovery and discovery of microlensing events are possible with a careful consideration of photometric systematics. This work can help inform predictions about the observability of microlensing signals in future wide-field time-domain surveys such as that of LSST.

@@

\section{Introduction}
Since Einstein first studied gravitational microlensing (cite), the phenomena has been used to study dark objects, probe galactic structure and kinematics, and discover extrasolar planets. Microlensing events were once relatively rare, but with advances in CCD technology and the help of several dedicated microlensing surveys (), $>500$ lensing events are now detected per observing season.

[MORE MICROLENSING INTRODUCTION]

[QUICK INTRO TO THEORY AND JARGON]

@@

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis (cite Lupton?) to identify transient events in raw imaging data. Light curves of any transient sources are then analyzed and further pruned using a variety of selection methods to search for microlensing event candidates (cite MACHO, OGLE). Using difference imaging is essential for studying crowded fields such as the Galactic bulge where photometry is difficult (cite or figure?), and also limits the number of light curves that must be searched for microlensing events, thus saving on computation time. Pre-selection with difference imaging also enables fast, real-time detection of events, allowing dedicated microlensing surveys to obtain high-cadence follow-up observations of interesting sources. Such surveys also have the advantage of relatively uniform time sampling of the survey footprint over an observing season.

@@

Previous work has used a variety of survey-specific selection methods (as described in \citealt{alcock2000, wyrzykowski2009, hamadache2009, sumi2011}), but the general idea is to \begin{enumerate}
	\item detect transient events using difference image analysis, 
	\item require that any selected light curves have some number of consecutive points brighter than some threshold,
	\item require that a microlensing model fit best describes the data using some cut on $\chi^2$,
	\item require that the model event parameters have reasonable values.
\end{enumerate}
Applying the second cut above to irregularly sampled data is more challenging and more computationally expensive because we must account for gaps in our data; it isn't obvious that this same selection procedure will be as successful for the PTF data. Motivated by work on identifying and classifying general variability in light curve data \citep{shin2009}[others?], we set out to compare the \textit{detection efficiency} of performing microlensing event selection with a set of variability statistics to previously used methods.

\subsection{Event Selection} \label{sec:event_selection}
We choose five statistical measures of variability, as compiled by \cite{shin2009}, that have been previously applied to classification of periodic variables, and evaluate their effectiveness in identifying microlensing events. The statistics, or \emph{variability indices}, are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$ (we choose not to implement the sixth, $AoVM$, because it mainly helps with periodic sources). $\sigma/\mu$ is the ratio of the root-variance to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations in the light curve. We modify their definition of $Con$ such that it is the number of clusters of three or more consecutive observations that are more than $3\sigma$ brighter than the reference magnitude of the source (e.g., for a single microlensing event in an otherwise flat light curve this value should be $Con=1$). This change allows us to use the performance of $Con$ as a measure of Cut 2 (\S\ref{sec:event_recovery}). $\eta$ is the von Neumann ratio, originally proposed in \cite{von_neumann1941}, and is a ratio of the mean square successive difference over the sample variance:
\begin{align}
	\eta = \frac{\delta^2}{\sigma^2} = \frac{\sum^{N-1}_i(x_{i+1} - x_i)^2/(N-1)}{\sigma^2}
\end{align}
$\eta$ is small, but still greater than zero, when there is strong serial correlation between successive data points. $J$ and $K$ were suggested by \cite{stetson1996}:
\begin{align}
	\delta_i &= \sqrt{\frac{N}{N-1}}\frac{x_i-\mu}{e_i}\\
	J &= \sum^{N-1}_i sign(\delta_i \delta_{i+1})\sqrt{|\delta_i \delta_{i+1}|}\\
	K &= \frac{1/N\sum^N_i |\delta_i|}{\sqrt{1/N\sum^N_i\delta_i^2}}
\end{align}
where $e_i$ is the photometric error of each data point. $J$ tends to 0 for non-variable stars, but is large when there are significant differences between successive data points in a light curve. $K$ is a measure of the kurtosis or ``peakedness'' of the distribution of points in the light curve. We add one more variability index motivated by Cut 3 (\S\ref{sec:event_recovery}): $\Delta \chi^2$ is defined as the difference in chi-squared between fitting a five-parameter Gaussian model and fitting a linear model to the light curve.

@@

\subsection{Detection Efficiency}
The probability of detecting any given microlensing event depends on the parameters of the event, the survey properties (observing strategy, limiting magnitude, weather), and the nature of the stellar field (crowding). To compute the detection efficiency, $\varepsilon$, for a given selection criterium, we run Monte Carlo simulations to artificially add microlensing events to real data and evaluate how often these events are recovered by the cut. This analysis must be done on a field-by-field basis, however we can compare our selection methods on a sub-sample of representative cases.

For a particular field, we randomly sample 1,000 light curves from each chip with $>25$ ``good'' observations (good observations are those that have no IPAC or SExtractor flags set [is there some IPAC paper to cite here?]). With this sample, we compute the set of variability statistics (\S\ref{sec:event_selection}), then iterate to add different simulated microlensing events to each light curve. We use a random number generator to only add events during $\apwsim 50\%$ of the iterations, which allows us to later quantify the false positive rate of our selection methods. For each iteration we recompute the variability statistics and store the event parameters. For each simulated event, the peak time, $t_0$, is drawn from a uniform distribution between the minimum and maximum observation times, the impact parameter, $u_0$, is drawn from a uniform distribution between 0 and 1, and the event timescale, $t_E$, is drawn from a log-uniform distribution between 1 and 1000 days.

We apply our selection criteria to each simulated light curve using the variability statistics computed for the initial light curves (from \textit{before} we add simulated events), and then bin the selected data by the event timescale.

- Previous surveys had fairly uniform time sampling, and smaller areas

- With PTF, fields have vastly different sampling patterns, baselines, and cadences

- We have to consider the efficiency on a field-by-field basis

- Show detection efficiency of each variability statistic

@@

@@

\begin{figure}
	\centering
	\caption{A visualization of the different sampling regimes of six different PTF fields over the same one year baseline. Darker points mean more data (more exposures). Some fields have roughly uniform coverage, where others are very dense during some weeks and not others. }
    \includegraphics[width=1.0\textwidth]{figures/sampling_figure.pdf}
    \label{fig:sampling}
\end{figure}
%
%\begin{figure}
%	\centering
%	\caption{Example of how the variability indices may be used to separate microlensing events from periodic or flat light curves.}
%    \includegraphics[width=1.0\textwidth]{figures/J_vs_K_Con_figure.png}
%    \label{fig:var_idx1}
%\end{figure}
%
%\begin{figure}[htbp]
%	\centering
%	\caption{Two neighboring light curves. The time scale is the same on both plots. Note the strong similarities in the overall shapes of the data.} 
%	 \includegraphics[width=1.0\textwidth]{figures/bad_data_figure.pdf}
%	\label{fig:lc_correlated}
%\end{figure}

>>>>

\begin{abstract}
[APW: update the abstract..]
Many current photometric, time-domain surveys are driven by specific goals, such as supernova searches, transiting exoplanet discoveries, or stellar variability studies, which set the cadence with which individual fields get re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to an extremely non-uniform sampling gradient over the survey footprint of nearly 20,000 deg$^2$: while the typical 7.26~deg$^2$ PTF field has been imaged 15 times, \apwsim1000~deg$^2$ of the survey has been observed more than 150 times. We use the existing PTF data to study the trade-off between a large survey footprint and irregular sampling when searching for microlensing events, and to examine the probability that such events can be recovered in these data. We conduct Monte Carlo simulations to evaluate our detection efficiency in a hypothetical survey field as a function of both the baseline and number of observations. We also apply variability statistics to systematically differentiate between periodic, transient, and flat light curves. Preliminary results suggest that both recovery and discovery of microlensing events are possible with a careful consideration of photometric systematics. This work can help inform predictions about the observability of microlensing signals in future wide-field time-domain surveys such as that of LSST.

@@

\section{Introduction}

\subsection{Overview}
General relativity tells us that mass alters the propagation of light. In particular, light from a background source passing some mass distribution will be deflected by an angle that depends on the distance between the source and mass distribution, as well as the structure and kinematics of the mass distribution itself. For an observer aligned with a source and a point lensing mass, the relevant angular scale for the observed deflection is given by the Einstein ring radius:
\begin{equation} \label{eq:einstein_radius}
	\theta_E = \sqrt{\frac{4GM}{c^2}\frac{D_{S} - D_L}{D_L D_S}}
\end{equation}
where $D_L$ is the distance to the lensing object, $D_S$ is the distance to the source, and $M$ is the mass of the lens.

Microlensing designates any unresolvable gravitational lensing, where the images formed by the lens just contribute to an increase in flux. For Galactic distance and mass scales, the Einstein radius (Eq.~\ref{eq:einstein_radius}) can be expressed as:
\begin{align}
	\theta_E &\approx 550\mathrm{\mu as}~\sqrt{\frac{1-D_L/D_S}{D_L/D_S}}~\Big(\frac{D_S}{8~\mathrm{kpc}}\Big)^{-1/2} \Big(\frac{M}{0.3~M_\sun}\Big)^{1/2}
\end{align}
\citep{paczynski1986, gaudi2011}. A solar-mass star in the bulge lensing another star in the bulge would have an Einstein radius of about $\theta_E \sim 0.3~\mathrm{mas}$, well below the resolution limit of even our most powerful telescopes.

Since Einstein first mentioned gravitational microlensing \citep{einstein1936}, it has been applied to the search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, galactic structure and kinematics \citep{binney2000}, and extrasolar planets \citep{}[APW: cite]. Microlensing events were once a relatively rare phenomena, but with advances in CCD technology and the help of several dedicated microlensing surveys [APW: cite], more than [APW: check number]500 events are detected each year by several microlensing collaborations and surveys. Within the next decade, this number is expected to [] as the next generation of photometric, time-domain surveys come online [APW: cite LCOGT, LSST, ??].

[APW: Some brief description of what is PTF, why it is the bridge between what has been done and what will be done, etc...]

\subsection{Something??}
Assuming that the source is an un-blended point source and the lens is a foreground, dim object, a microlensing event is fully described by three parameters: the angular impact parameter $u_0$, the peak time of the event $t_0$, and the timescale of the event (Einstein crossing time) $t_E$ such that the increase in flux due to source magnification is given by:
\begin{align}
	F(t) &= A(t)\times F_{source} \\
	A(u) &= \frac{u^2 + 2}{u\sqrt{u^2 + 4}}\\
	u(t) &= \sqrt{u_0^2 + 2\Big(\frac{t-t_0}{t_E}\Big)}
\end{align}
where $A(t)$ is the amplification factor.

The microlensing perturbation can also be expressed in terms of magnitudes as:
\begin{align}
	m(t) &= m_0(t) - 2.5\log A(t)
\end{align}
where $m_0(t)$ is the unperturbed, possibly time-variable magnitude of the source.

%However, if the event is \textit{too} bright, the pixels will saturate and there could be photometric errors flagging the most interesting data as bad.
When $u_0$ is small ($u_0<<1$), the amplification is large. If $t_E$ is short ($t_E \lesssim\mathrm{days}$), surveys may miss or poorly sample the event, but if $t_E$ is long ($t_E \gtrsim 1~\mathrm{year}$) the event may be confused with other forms of long-duration variability. \figurename~\ref{fig:microlensing_sim} shows a ``typical'' PTF light curve and demonstrates what microlensing events with $t_E=20~\mathrm{days}$ and varying $u_0$ may look like in the data.

The microlensing event rate in some particular region of sky depends on the space and velocity distributions of sources and on the space, velocity, and mass distribution of lenses. To maximize the event rate, microlensing surveys have typically focused their observations towards high stellar density regions such as the Galactic Bulge, M31, and the Magellanic clouds. Microlensing events do occur away from stellar population centers, but without a large-area, time-domain search of these regions, only one such event has been recorded ([APW: cite Fukui 2007, Gaudi 2008 RE: Tago event). [APW] Gaudi 2008 were able to establish probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the location of the event. With JHK magnitudes of the lens itself or a direct measurement of the proper velocity of the lens, these limits could be turned into precise measurements. Such a detailed analysis is not normally possible for events in the previously mentioned high-density regions where the lenses are much farther away, and blending makes precise photometry difficult.

Not all events away from the bulge will be as fortunate as the Tago event, but lensing events away from stellar population centers are generally expected to have closer sources, closer lenses, and less crowded background fields compared to  those found by modern microlensing surveys ([APW: cite Di Stefano mesolensing]). Thus, the properties of the lenses are more likely to be well-constrained by direct observation, possibly enabling mass measurements of objects not bound in binary pairs. The event rate in low stellar density regions will certainly be smaller than the event rate near the Bulge \citep[e.g.,][]{wood_optical_depth, ogle_optical_depth, macho_optical_depth, eros_optical_depth}, but the opportunity for precise, direct observations of lenses would be extremely valuable for studying low-mass stars, substellar objects, and stellar remnants such as isolated neutron stars [APW: cite NS work?].

@@

\section{Microlensing event recovery} \label{sec:event_recovery}
Microlensing surveys typically use difference image analysis (cite Lupton?) to identify transient events in raw imaging data. Light curves of any transient sources are then analyzed and further pruned using a variety of selection methods to search for microlensing event candidates (cite MACHO, OGLE). Using difference imaging is essential for studying crowded fields such as the Galactic bulge where photometry is difficult (cite or figure?), and also limits the number of light curves that must be analyzed, thus saving on computation time. Pre-selection with difference imaging also enables fast, real-time detection of events, allowing dedicated microlensing surveys to obtain high-cadence follow-up observations of interesting sources. Such surveys also have the advantage of relatively uniform time sampling of the survey footprint over an observing season.

@@

Previous work has used a variety of survey-specific selection methods (as described in \citealt{alcock2000, wyrzykowski2009, hamadache2009, sumi2011}), but the general idea is to \begin{enumerate}
	\item detect transient events using difference image analysis, 
	\item require that any selected light curves have some number of consecutive datapoints brighter than some threshold,
	\item require that a microlensing model fit best describes the data using some cut on $\chi^2$,
	\item require that the model event parameters have reasonable values.
\end{enumerate}
Applying the second cut above to irregularly sampled data is more challenging and more computationally expensive because we must account for gaps in our data. Furthermore, it isn't obvious that this same selection procedure will be as successful for the PTF data. Motivated by work on identifying and classifying general variability in light curve data \citep{shin2009}[others?], we set out to compare the performance of microlensing event selection using a set of variability statistics as an initial cut in our candidate detection pipeline.

\subsection{Event Selection} \label{sec:event_selection}
We choose five statistical measures of variability, as compiled by \cite{shin2009}, that have been previously applied to classification of periodic variables, and evaluate their effectiveness in recovering simulated microlensing events in the data. These statistics, or \emph{variability indices}, are $\sigma/\mu$, $Con$, $\eta$, $J$, and $K$ (we choose not to implement the sixth, $AoVM$, because it mainly helps with periodic sources). $\sigma/\mu$ is the ratio of the root-variance to the sample mean, 
\begin{align}
	\frac{\sigma}{\mu} = \frac{\sqrt{\sum^N_i (x_i - \mu)^2 / (N-1)}}{\sum^N_i x_i/N},
\end{align}
where N is the total number of observations in the light curve. We modify their definition of $Con$ such that it is the number of clusters of three or more consecutive observations that are more than $3\sigma$ brighter than the reference magnitude of the source (e.g., for a single microlensing event in an otherwise flat light curve this value should be $Con=1$). This change allows us to use the performance of $Con$ as a proxy for Cut 2 (\S\ref{sec:event_recovery}). $\eta$ is the von Neumann ratio, originally proposed in \cite{von_neumann1941}, and is a ratio of the mean square successive difference over the sample variance:
\begin{align}
	\eta = \frac{\delta^2}{\sigma^2} = \frac{\sum^{N-1}_i(x_{i+1} - x_i)^2/(N-1)}{\sigma^2}
\end{align}
$\eta$ is small, but still greater than zero, when there is strong serial correlation between successive data points. $J$ and $K$ were suggested by \cite{stetson1996}:
\begin{align}
	\delta_i &= \sqrt{\frac{N}{N-1}}\frac{x_i-\mu}{e_i}\\
	J &= \sum^{N-1}_i sign(\delta_i \delta_{i+1})\sqrt{|\delta_i \delta_{i+1}|}\\
	K &= \frac{1/N\sum^N_i |\delta_i|}{\sqrt{1/N\sum^N_i\delta_i^2}}
\end{align}
where $e_i$ is the photometric error of each data point. $J$ tends to 0 for non-variable stars, but is large when there are significant differences between successive data points in a light curve. $K$ is a measure of the kurtosis or ``peakedness'' of the distribution of points in the light curve. We add one more variability index motivated by Cut 3 (\S\ref{sec:event_recovery}): $\Delta \chi^2$ is defined as the difference in chi-squared between fitting a five-parameter Gaussian model and fitting a linear model to the light curve. \figurename~\ref{fig:indices_examples} shows examples of a median light curve for each variability statistic (left panels) along with the most extreme outliers (right panels) for XXX light curves on a single CCD on PTF field YYY.

\figurename~\ref{fig:scatter_plot_matrix} shows a scatter-plot matrix of all six of our variability statistics. In this figure, we take 10,000 light curves from a single PTF field (field id = ), compute the variability indices for each light curve, and then plot $1\sigma$ (red) and $2\sigma$ (green) contours in each panel for the resultant distributions. We then add 10 simulated microlensing events to each light curve, recompute the variability indices, and plot these values as black points.

@@

\subsection{Detection Efficiency}
The probability of detecting any given microlensing event depends on the parameters of the event, the survey properties (observing strategy, limiting magnitude, weather), and the nature of the stellar field (crowding). To compute the detection efficiency, $\varepsilon$, for a given selection criterium, we run Monte Carlo simulations to artificially add microlensing events to real data and evaluate how often these events are recovered by the cut. This analysis must be done on a field-by-field basis, however we can compare our selection methods on a sub-sample of representative cases. Performing these simulations with real data allows us to compute the ... [APW: explain difference between two different detection efficiencies (of selection method vs. of survey parameters), and how we marginalize over both].

For a particular field, we randomly sample 1,000 light curves from each chip with $>25$ ``good'' observations (good observations are those that have no IPAC or SExtractor flags set [Marcel: is there some IPAC paper to cite here?]). With this sample, we compute the set of variability statistics (\S\ref{sec:event_selection}), then iterate to add [APW: N] different simulated microlensing events to each light curve. We use a random number generator to only add events during $\apwsim 50\%$ of the iterations, which allows us to later quantify the false positive rate of our selection methods. For each iteration we recompute the variability statistics and store the event parameters. [The parameters for each simulated event, the peak time, $t_0$, is drawn from a uniform distribution between the minimum and maximum observation times, the impact parameter, $u_0$, is drawn from a uniform distribution between 0 and 1, and the event timescale, $t_E$, is drawn from a log-uniform distribution between 1 and 1000 days.]

We apply our selection criteria to each simulated light curve using the variability statistics computed for the initial light curves (from \textit{before} we add simulated events), and then bin the selected data by the event timescale [APW: figure reference].

@@

\begin{figure}
	\centering
	\caption{(a) the light curve plotted as is, (b) light curve with a simulated microlensing event with $u_0=1.0$, $t_E=20~\mathrm{days}$, (c) same as (b) with $u_0=0.5$, (d) same as (b) with $u_0=0.01$.}
    \includegraphics[width=1.0\textwidth]{figures/simulated_events.pdf}
    \label{fig:microlensing_sim}
\end{figure}

@@

\begin{figure}
	\centering
	\caption{A visualization of the different sampling regimes of six different PTF fields over the same one year baseline. Darker points mean more data (more exposures). Some fields have roughly uniform coverage, where others are very dense during some weeks and not others. }
    \includegraphics[width=1.0\textwidth]{figures/sampling_figure.pdf}
    \label{fig:sampling}
\end{figure}

%
%\begin{figure}
%	\centering
%	\caption{Example of how the variability indices may be used to separate microlensing events from periodic or flat light curves.}
%    \includegraphics[width=1.0\textwidth]{figures/J_vs_K_Con_figure.png}
%    \label{fig:var_idx1}
%\end{figure}
%
%\begin{figure}[htbp]
%	\centering
%	\caption{Two neighboring light curves. The time scale is the same on both plots. Note the strong similarities in the overall shapes of the data.} 
%	 \includegraphics[width=1.0\textwidth]{figures/bad_data_figure.pdf}
%	\label{fig:lc_correlated}
%\end{figure}
